---
title: "Processing a WHO air quality data file"
output:
  pdf_document: default
  html_notebook: default
---

Load the [Excel spreadsheet](https://www.who.int/phe/health_topics/outdoorair/databases/who-aap-database-may2016.xlsx) from the [WHO Global Urban Ambient Air Pollution Database](https://www.who.int/phe/health_topics/outdoorair/databases/cities/en)

```{r}
# Load the packages for reading Excel files
# Use the Bristol R package mirror
bristol_cran <- 'https://www.stats.bris.ac.uk/R'
install.packages('rJava', repo=bristol_cran)
install.packages("XLConnect", repo=bristol_cran)
library(XLConnect)
```

Load the relevant worksheet from the Excel file:

```{r}
workbook <- loadWorkbook('aap_air_quality_database_2018_v14.xlsx')
data <- XLConnect::readWorksheet(workbook, 'database', startRow=3)
head(data)
```

```{r}
str(data)
```

We are going to use some [Tidyverse](https://www.tidyverse.org) packages to organize the data.

```{r}
install.packages("tidyverse", repo=bristol_cran)
```

`dplyr` is part of the Tidyverse.  It has some nice utilities for selecting and renaming columns in a data frame.  You can so this without `dplyr`, but `dplyr` makes it much easier to express in a few lines of code.

```{r}
library(dplyr)
```

Notice there are two sets of columns starting `Annual.mean`.

```{r}
str(select(data, starts_with('Annual.mean')))
```

Likewise for `Temporal.coverage`:

```{r}
str(select(data, starts_with('Temporal.coverage')))
```

Looking at the column names, and at the original Excel file, the first in each pair refers to PM10, and the second to PM2.5 : https://en.wikipedia.org/wiki/Particulates

We rename the columns accordingly:

```{r}
data <- rename(data,
               'pm10_yr_mean'='Annual.mean..ug.m3',
               'pm10_temp_cover'='Temporal.coverage',
               'pm25_yr_mean'='Annual.mean..ug.m3.1',
               'pm25_temp_cover'='Temporal.coverage.1'
)
head(data)
```

Looking through the data (perhaps with the `View` function), it looks like the particulate concentration columns have form `(<number>)-converted value` if the value is converted.

```{r}
head(data$pm25_yr_mean)
```

First we check that the `note.on.converted.PM25` column always says `Converted` when the PM2.5 concentration column has `-converted value`:

```{r}
is_converted <- data$note.on.converted.PM2.5 == 'Converted'
head(is_converted)
```

Do all the `Converted` values have `-converted value` in their concentration column?  Do `Measured` values never have `-converted value` in their concentration column?

Here we use the `stringr` package to make it easier to work with strings.  `stringr` is one of the packages in the Tidyverse.

```{r}
library(stringr)
```

For each value in the concentration column, we test whether it contains the string `-converted value`.

```{r}
has_converted_value = str_detect(data$pm25_yr_mean, '-converted value')
head(has_converted_value)
```

We confirm that there is complete equivalence between the `Converted` values in the `note.on.converted.PM2.5` column, and the presence of `-converted value` in the concentration column.

```{r}
all(is_converted == has_converted_value)
```

Do the same for the PM10 columns:

```{r}
is_conv_pm10 <- data$note.on.converted.PM10 == 'Converted'
all(is_conv_pm10 == str_detect(data$pm10_yr_mean, '-converted value'))
```

The equivalence we found means that we can throw away the `converted value` part of the concentration values, because the information is already in the matching `note.on.converted.PM` column.

Now we need to convert the concentration column into numbers, by removing the `-converted value` part, and converting to numeric values.  Because we will do this for two columns, we put the logic into a function, so we can repeat it, without retyping.

```{r}
recode_col <- function(col) {
  # Recode column with "converted" values as numeric
  # Get number out of '(number)-converted value', if present
  conv_strs <- str_match(col, '\\((\\d+)\\)-converted value')[, 2]
  # If value did not match '(number)-converted value', we get NA
  not_converted = is.na(conv_strs)
  conv_values <- numeric(length(col))
  conv_values[not_converted] = as.numeric(col[not_converted])
  converted = !not_converted
  conv_values[converted] = as.numeric(conv_strs[converted])
  conv_values
}
```

Apply the function to the two concentration columns:

```{r}
data$pm10_yr_mean = recode_col(data$pm10_yr_mean)
data$pm25_yr_mean = recode_col(data$pm25_yr_mean)
head(data)
```

Actually, there are a few typos in the `notes.on.converted.PM` fields:

```{r}
table(data$note.on.converted.PM2.5)
```

```{r}
table(data$note.on.converted.PM10)
```

We need to set `measured` to be `Measured` and `converted` to be `Converted`.  To do this, we use the `str_to_title` function from the `stringr` library that we have already loaded:

It works like this:

```{r}
str_to_title('measured')
```

We apply it to the whole vector, for both columns:

```{r}
data$note.on.converted.PM2.5 <- str_to_title(data$note.on.converted.PM2.5)
table(data$note.on.converted.PM2.5)
data$note.on.converted.PM10 <- str_to_title(data$note.on.converted.PM10)
table(data$note.on.converted.PM10)
```

Next we process the `Region` field.  It consists of two pieces of information: the region name, and the income label, that can be `HIC` (High Income Country) or `LMIC` (Lower Middle Income Country):

```{r}
region_values <- data$Region
head(region_values)
```

Time for some more fancy string manipulation:

```{r}
# Split the string into the region name and the income catagory.
regions_types <- str_match(region_values, '(.*)\\s+\\((.*)\\)')
head(regions_types)
```

Replace original `Region` column with region name, and add new `income_category` column:

```{r}
data$Region <- regions_types[, 2]
data$income_category <- regions_types[, 3]
head(data)
```

Confirm the number of income categories:

```{r}
table(data$income_category)
```

For convenience, we would like `income_category` to come after `Region` in the sequence of columns.

```{r}
col_names <- names(data)
col_names
```

```{r}
# Reorder, to put income category after Region
new_names = c(col_names[1], 'income_category', col_names[2:15])
new_names
```

Reorder data frame with new column order.

```{r}
data <- select(data, new_names)
str(data)
```

We need a column for unique city / town name.  For example, there is more than one `London`, meaning the `City.Town` values are not unique to a city or town:

```{r}
town_is_london <- data$City.Town == 'London'
subset(data, town_is_london)
```

Put the country name into the `City.Town` name, to disambiguate.

```{r}
data$City.Town <- paste(data$City.Town, ' (', data$Country, ')', sep='')
head(subset(data, town_is_london))
```

Write the converted data frame to a CSV file.

```{r}
write.csv(data, 'aap_air_quality_database_2018_v14.csv', row.names=FALSE)
```

Check we can load the CSV file back into memory.

```{r}
csv_data <- read.csv('aap_air_quality_database_2018_v14.csv')
head(csv_data)
```
